# Markov Chain Monte Carlo

## Eample: Normal mixture
Generate some data with mixing rate $\delta$.
```{r nmix-sim}
delta <- 0.7 # true value to be estimated based on the data
n <- 100
set.seed(254)
u <- rbinom(n, prob = delta, size = 1)
x <- rnorm(n, ifelse(u == 1, 7, 10), 0.5)
```

Assuming that the normal distribution parameters $\mu_1$, $\sigma_1^2$, $\mu_2$, and $\sigma_2^2$ are known. The only unknown parameter is $\delta$. The likelihood function of $\delta$ is
```{r nmix-like}
mylike <- function(delta, x) {
    prod(delta * dnorm(x, 7, 0.5) + (1 - delta) * dnorm(x, 10, 0.5))
}
```


We impose an uninformative prior on $\delta$ and use a simple random walk proposal to construct the Metropolis--Hasting algorithm.
```{r nmix-mcmc}
## simple random walk chain
myRange <- function(v, width) {
    min(1, v + width) - max(0, v - width)
}

mymcmc <- function(niter, init, x, width) {
    v <- double(niter)
    for (i in 1:niter) {
        cand <- runif(1, max(0, init - width), min(1, init + width))
        ratio <- mylike(cand, x) / myRange(cand, width) /
            mylike(init, x) * myRange(init, width)
        if (runif(1) < min(ratio, 1)) {
            v[i] <- init <- cand
        } else v[i] <- init
    }
    v
}
```


With initial value $\delta_0 = .2$ and stepsize $.1$ in the random walk proposal, we run the 2500 iterations and throw away the first 500 iterations.
```{r nmix-res}
niter <- 2500
plot(ts(z <- mymcmc(niter, .2, x, .1)[-(1:500)]))
hist(z)
```

## Exercises

### Normal mixture revisited
Consider again the normal mixture example, except that the parameters of the normal distributions are considered unknown. Suppose that prior for $\mu_1$ and $\mu_2$ are $N(0, 10^2)$, that the prior for $1/\sigma_1^2$ and $1/\sigma_2^2$ are $\Gamma(a, b)$ with shape $a = .5$ and scale $ b = 10$. Further, all the priors are independent. Design an MCMC using the Gibbs sampling approach to estimate all 5 parameters. You may use the __arms()__ function in package **HI**. Run your chain for sufficiently long and drop the burn-in period. Plot the histogram of the results for all the parameters.
