# Markov Chain Monte Carlo

## Eample: Normal mixture
Generate some data with mixing rate $\delta$.
```{r nmix-sim}
delta <- 0.7 # true value to be estimated based on the data
n <- 100
set.seed(123)
u <- rbinom(n, prob = delta, size = 1)
x <- rnorm(n, ifelse(u == 1, 7, 10), 0.5)
```

Assuming that the normal distribution parameters $\mu_1$, $\sigma_1^2$, $\mu_2$, and $\sigma_2^2$ are known. The only unknown parameter is $\delta$. The likelihood function of $\delta$ is
```{r nmix-like}
mylike <- function(delta, x) {
    prod(delta * dnorm(x, 7, 0.5) + (1 - delta) * dnorm(x, 10, 0.5))
}
```


We impose an uninformative prior on $\delta$ and use a simple random walk proposal to construct the Metropolis--Hasting algorithm.
```{r nmix-mcmc}
## simple random walk chain
myRange <- function(v, width) {
    min(1, v + width) - max(0, v - width)
}

mymcmc <- function(niter, init, x, width) {
    v <- double(niter)
    for (i in 1:niter) {
        cand <- runif(1, max(0, init - width), min(1, init + width))
        ratio <- mylike(cand, x) / myRange(cand, width) /
            mylike(init, x) * myRange(init, width)
        if (runif(1) < min(ratio, 1)) {
            v[i] <- init <- cand
        } else v[i] <- init
    }
    v
}
```


With initial value $\delta_0 = .2$ and stepsize $.1$ in the random walk proposal, we run the 2500 iterations and throw away the first 500 iterations.
```{r nmix-res}
niter <- 2500
plot(ts(z <- mymcmc(niter, .2, x, .1)[-(1:500)]))
hist(z)
```


## General-Purpose Gibbs Sampling with the ARMS

Consider an example of posterior inference. 
Suppose that $(Y_i, X_i)$, $i = 1, \ldots, n$, are observed, where
\[
Y_i \mid X_i \sim \mbox{Poisson}(\exp(a + b X_i)).
\]
This is a Poisson regression model. Assume independent prior distributions on $a$ and $b$ with $N(0, \sigma^2)$ and $N(0, \tau^2)$. The posterior density of $(a, b)$ is
\[
q(a, b \mid \mbox{data}) \propto
\exp\left(a \sum_i^n Y_i + b \sum_i^n X_i Y_i - e^a \sum_i^n e^{X_i b} - \frac{a^2}{2\sigma^2} - \frac{b^2}{2\tau^2}\right).
\]
The full conditional distributions of $a$ and $b$ can be shown to be log-concave, which allows adaptive rejection alogorithm [@robert2004monte]. One could also just use the general ARMS as a lazy man's approach.


Let us first generate some data.
```{r pois-gendata}
n <- 100
a <- 0.0; b <- 0.5
x <- rnorm(n)
y <- rpois(n, exp(a + b * x))
mydata <- data.frame(y = y, x = x)
```
The posterior density up to an unknown normalizing constant can be easily computed.
```{r pois-post}
logpost <- function(theta, data, sigma2, tau2) {
  a <- theta[1]; b <- theta[2]
  x <- data$x; y <- data$y
  return(a * sum(y) + b * sum(x * y) - exp(a) * sum(exp(b * x))
         - a^2 / 2 / sigma2 - b^2 / 2 / tau2)
}
```
An MCMC based the Gibbs sampler uses the ARMS algorithm from R package __HI__.
```{r pois-mcmc}
mymcmc <- function(niter, thetaInit, data, sigma2, tau2) {
  p <- length(thetaInit)
  thetaCurrent <- thetaInit
  out <- matrix(NA, niter, p)
  for (i in 1:niter) {
    for (j in 1:p) {
      logFC <- function(thj) {
        theta <- thetaCurrent
        theta[j] <- thj
        logpost(theta, data, sigma2, tau2)
      }
      ## arms algorithm
      out[i, j] <- thetaCurrent[j] <-
        HI::arms(thetaCurrent[j], logFC, 
                 function(x) ((x > -10) * (x < 10)), 1)
    }
  }
  out
}
```
Now give it a try with $\sigma^2 = 100$ and $\tau^2 = 100$.
```{r rpois-run}
niter <- 500
thetaInit <- c(2, 2)
sigma2 <- tau2 <- 100
sim <- mymcmc(niter, thetaInit, mydata, sigma2, tau2)
plot(ts(sim[,1]))
plot(ts(sim[,2]))
```
## Exercises

### Normal mixture revisited
Consider again the normal mixture example, except that the parameters of the normal distributions are considered unknown. Suppose that prior for $\mu_1$ and $\mu_2$ are $N(0, 10^2)$, that the prior for $1/\sigma_1^2$ and $1/\sigma_2^2$ are $\Gamma(a, b)$ with shape $a = .5$ and scale $ b = 10$. Further, all the priors are independent. Design an MCMC using the Gibbs sampling approach to estimate all 5 parameters. You may use the __arms()__ function in package **HI**. Run your chain for sufficiently long and drop the burn-in period. Plot the histogram of the results for all the parameters.
