--- 
title: "STAT 5361: Statistical Computing, Fall 2018"
author: "Jun Yan"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: chicago
link-citations: yes
description: "This is a series of notes
for the students of STAT 5361,
Statisticl Computing, at UConn."
---

# Prerequisites

We assume that students can use R/Rstudio comfortably. If you are new to R, an extra amount of hard work should be devoted to make it up. In addition to the official documentations from the R-project, the presentations at a [local SIAM workshop in Spring 2018] (https://siam.math.uconn.edu/events/) given by Wenjie Wang, a PhD student in Statistics at the time, can be enlighting:

- <https://github.com/wenjie2wang/2018-01-19-siam>
- <https://github.com/wenjie2wang/2018-04-06-siam>

If you have used R, but never paid attention to R programming styles, a style clinic would be a necessary step. A good place to start is Google's R style guide at <https://google.github.io/styleguide/Rguide.xml>. From my experience, the two most commonly overlooked styles for beginners are spacing and indentation. Appropriate spacing and indentation would immediately make crowdly piled code much more eye-friendly. Such styles can be automatically enforced by R packages such as **formatr** or **lintr**. Tow important styles that cannot be automatically corrected are naming and documentation. As in any programming languate, naming of R objects (variables, functions, files, etc.) shoud be informative, concise, and consistent with certain naming convention. Documentation needs to be sufficient, concise, and kept close to the code; tools like R package **roxygen2** can be very helpful. See Hadley Wichham's online book <http://style.tidyverse.org/> for more detailed tips.


For intermediate R users who want a skill lift, The __Advanced R Programming__ book by Hadley Wickham is available at <https://adv-r.hadley.nz/>. The source that generated the book is kindly made available at GitHub: <https://github.com/hadley/adv-r>. It is a great learning experience to complile the book from the source, during which you may pick up many necessary skills.


The homework, exam, and project will be completed by __R Markdown__. Following the step by step the instructions in Yihui Xie's online book on __bookdown__ at <https://bookdown.org/yihui/bookdown/>, you will be amazed how quickly you can learn to produce cool-looking documents and even book manuscripts. If you are a keener, you may as well following Yihui's __blogdown__, see online book <https://bookdown.org/yihui/blogdown/>, to build your own website using R Markdown.


All your source code will be version controlled by __git__ and archived on __GitHub__. RStudio has made using git quite straightforward. The online tutorial by Jenny Bryan, Happy Git and GitHub for the useR, is a very useful tool to get started

## A Teaser Example: Likelihood Estimation

In mathematical statistics, we have learned that, under certain regularity conditions, the maximum likelihood estimator (MLE) is consistent, asymptotically normal, and most efficient. The asymptotic variance of the estimator if the inverse of the Fisher information matrix. Specifically, let $X_1, \ldots, X_n$ be a random sample from a distribution with density $f(x; \theta)$, where $\theta$ is a parameter. How do we obtain the MLE?

```{r data}
set.seed(123)
n <- 100
x <- rgamma(n, shape = 2, scale = 4)
hist(x)
```

Package **MASS** provides a function `fitdistr()` to obtain the MLE for univariate distributions with a random sample. We can learn two things from this function. First, an objective function representing the negative loglikelihood is formed, depending on the input of the density function, and fed to the optimizer function `optim`. Second, the variance estimator of the MLE is obtained by inverting the Hessian matrix of the objective function, which is an estimator of the Fisher information matrix. For commonly used distributions, starting values are not necessary. The function computes moment estimator and use them as starting values.
```{r fitdistr}
MASS::fitdistr(x, densfun = "gamma")
```

For distributions not in the provided list, the `densfun` needs to be a function returning a density evaluated at its first arguments, in which case, the `start` argument needs to be a named list to feed to the optimizer. For example, pretend that someone has a fancy distribution which is just a rename of the gamma distribution. Its density function is defined based on the gamma density `dgamma`.
```{r fancy}
dfancy <- function(x, shape, scale, log = FALSE) {
  dgamma(x, shape = shape, scale = scale, log = log)
}

suppressWarnings(MASS::fitdistr(x, densfun = dfancy, start = list(shape = 10, scale = 20)))
```


The **stats4**  package provides MLE using S4 classes.
```{r mle}
nll <- function(shape, scale) -sum(dfancy(x, shape, scale, TRUE))
suppressWarnings(fit <- stats4::mle(nll, start = list(shape = 10, scale = 10)))
stats4::summary(fit)
```

## Computer Arithmetics

How does a computer represent a number? Computers use switches, which has only two states, on and off, to represent numbers. That is, it uses the binary number system. With a finite (though large) number switches on a computer, not all real numbers are representable. Binary numbers are made of bits, each of which represent a switch. A byte is 8 bits, which represents one character.

### Integers
An integer on many computers has length 4-byte or 32-bit.
It is represented by the coefficients $x_i$ in
\[
\sum_{i=1}^{32} x_i 2^{i-1} - 2^{31},
\]
where $x_i$ is either 0 or 1.

```{r int}
imax <- .Machine$integer.max
imax
imax + 1L
log(imax, base = 2)

u <- 0L
b <- 1L
for (i in 1L:31L) {
  u <- u + b
  b <- b * 2L
}
u

1000L * 1000L * 1000L
1000L * 1000L * 1000L * 1000L
1000L * 1000L * 1000L * 1000
```

### Floating Point
A floating point number is represented by 
\[
(-1)^{x_0}\left(\sum_{i=1}^t x_i 2^{-i}\right) 2^{k}
\]
for $x_i \in \{0, 1\}$,
where $k$ is an integer called the exponent,
$x_0$ is the sign bit.
The fractional part $\sum_{i=1}^t x_i 2^{-i}$ is the significand. By convention (for unique representation), the exponent is chosen so that the fist digit of the significand is 1, unless it would result in the exponent being out of range.


A standard double precision representation uses 8 bytes or 64 bits: a sign bit, an 11 bit exponent, and 52 bits for the significand. With 11 bits there are $2^{11} = 2048$ possible values for the exponent, which are usually shifted so that the allowed values range from $-1022$ to 1024, again with one special sign value.


The IEEE standard for floating point arithmetic calls for basic arithmetic operations to be performed to higher precision, and then rounded to the nearest representable floating point number. 


Arithmetic underflow occurs where the result of a calculation is a smaller number than what the computer can actually represent. Arithmetic overflow occurs when a calculation produces a result that is greater in magnitude than what the computer can represent.

```{r float}
options(digits = 20)
.Machine$double.eps
.Machine$double.base ^ .Machine$double.ulp.digits
1 + .Machine$double.eps == 1
1 + .Machine$double.base ^ (.Machine$double.ulp.digits - 1L) == 1


## floating point traps
.1 == .3 / 3
.3 / 3

seq(0, 1, by = .1) == 0.3
seq(0, 1, by = .1)

unique(c(.3, .4 - .1, .5 - .2, .6 - .3, .7 - .4))
c(.3, .4 - .1, .5 - .2, .6 - .3, .7 - .4)

## largest possible significand
u <- 0
for (i in 1L:53) u <- u + 2^(-i)
u

## and the largest possible exponent--note that calculating 2^1024
## directly overflows
u * 2 * 2 ^ 1023

## next largest floating point number overflows
(u * 2 * 2 ^ 1023) * (1 + 1/2^52)

## smallest possible mantissa and smallest possible exponent
1 / 2^52 * 2^(-1022)
1 / 2^52 * 2^(-1022) / 2
```



## Exercises
1. Use git to clone the source of Hadley Wickham's **Advanced R Programming** from his GitHub repository to a local space on your own computer. Build the book using RStudio. During the building process, you may see error messages due to missing tools on your computer. Read the error messages carefully and fix them, until you get the book built. You may need to install some R packages, some fonts, some latex packages, and some building tools for R packages. On Windows, some codes for parallel computing may not work and need to be commented out.


2. Use **bookdown** or **rmarkdown** to produce a report for the following task. Consider approximation of the distribution function of $N(0, 1)$,
\begin{equation}
\Phi(t) = \int_{-\infty}^t \frac{1}{\sqrt{2\pi}} e^{-y^2 / 2} dy,
(\#eq:cdf)
\end{equation}
by the Monte Carlo methods:
\begin{equation}
\hat\Phi(t) = \frac{1}{n} \sum_{i=1}^n I(X_i \le t),
\end{equation}
where $X_i$'s are iid $N(0, 1)$ variables.
Experiment with the approximation at $n \in \{10^2, 10^3, 10^4\}$
at $t \in \{0.0, 0.67, 0.84, 1.28, 1.65, 2.32, 2.58, 3.09, 3.72\}$ to form a table.
The table should include the true value for comparison.
Further, repeat the experiment 100 times. Draw box plots of the bias at all $t$.
The report should look like a manuscript, with a title, an abstract, and multiple sections. It should contain at least one math equation, one table, one figure, and one chunk of R code. The template of our Data Science Lab can be helpful: <https://statds.org/template/>, the source of which is at <https://github.com/statds/dslab-templates>.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
