--- 
title: "STAT 5361: Statistical Computing, Fall 2018"
author: "Jun Yan"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: chicago
link-citations: yes
description: "This is a series of notes for the students of 
  STAT 5361, Statisticl Computing, at UConn."
---

# Prerequisites {#prelim}

We assume that students can use R/Rstudio comfortably. If you are new to R, an
extra amount of hard work should be devoted to make it up. In addition to the
official documentations from the R-project, the presentations at a [local SIAM
workshop in Spring 2018] (https://siam.math.uconn.edu/events/) given by Wenjie
Wang, a PhD student in Statistics at the time, can be enlighting:
- <https://github.com/wenjie2wang/2018-01-19-siam>
- <https://github.com/wenjie2wang/2018-04-06-siam>

If you have used R, but never paid attention to R programming styles, a style clinic would be a necessary step. A good place to start is Google's R style guide at <https://google.github.io/styleguide/Rguide.xml>. From my experience, the two most commonly overlooked styles for beginners are spacing and indentation. Appropriate spacing and indentation would immediately make crowdly piled code much more eye-friendly. Such styles can be automatically enforced by R packages such as **formatr** or **lintr**. Tow important styles that cannot be automatically corrected are naming and documentation. As in any programming languate, naming of R objects (variables, functions, files, etc.) shoud be informative, concise, and consistent with certain naming convention. Documentation needs to be sufficient, concise, and kept close to the code; tools like R package **roxygen2** can be very helpful. See Hadley Wichham's online book <http://style.tidyverse.org/> for more detailed tips.


For intermediate R users who want a skill lift, The __Advanced R Programming__ book [@wickham2014advanced] by Hadley Wickham is available at <https://adv-r.hadley.nz/>. The source that generated the book is kindly made available at GitHub: <https://github.com/hadley/adv-r>. It is a great learning experience to complile the book from the source, during which you may pick up many necessary skills.


The homework, exam, and project will be completed by __R Markdown__. Following the step by step the instructions in Yihui Xie's online book on __bookdown__ [@xie2016bookdown].at <https://bookdown.org/yihui/bookdown/>, you will be amazed how quickly you can learn to produce cool-looking documents and even book manuscripts. If you are a keener, you may as well following Yihui's __blogdown__, see online book <https://bookdown.org/yihui/blogdown/>, to build your own website using R Markdown.


All your source code will be version controlled by __git__ and archived on __GitHub__. RStudio has made using git quite straightforward. The online tutorial by Jenny Bryan, Happy Git and GitHub for the useR, is a very useful tool to get started

## A Teaser Example: Likelihood Estimation

In mathematical statistics, we have learned that, under certain regularity conditions, the maximum likelihood estimator (MLE) is consistent, asymptotically normal, and most efficient. The asymptotic variance of the estimator if the inverse of the Fisher information matrix. Specifically, let $X_1, \ldots, X_n$ be a random sample from a distribution with density $f(x; \theta)$, where $\theta$ is a parameter. How do we obtain the MLE?

```{r data}
set.seed(123)
n <- 100
x <- rgamma(n, shape = 2, scale = 4)
hist(x)
```

Package **MASS** provides a function `fitdistr()` to obtain the MLE for univariate distributions with a random sample. We can learn two things from this function. First, an objective function representing the negative loglikelihood is formed, depending on the input of the density function, and fed to the optimizer function `optim`. Second, the variance estimator of the MLE is obtained by inverting the Hessian matrix of the objective function, which is an estimator of the Fisher information matrix. For commonly used distributions, starting values are not necessary. The function computes moment estimator and use them as starting values.
```{r fitdistr}
MASS::fitdistr(x, densfun = "gamma")
```

For distributions not in the provided list, the `densfun` needs to be a function returning a density evaluated at its first arguments, in which case, the `start` argument needs to be a named list to feed to the optimizer. For example, pretend that someone has a fancy distribution which is just a rename of the gamma distribution. Its density function is defined based on the gamma density `dgamma`.
```{r fancy}
dfancy <- function(x, shape, scale, log = FALSE) {
  dgamma(x, shape = shape, scale = scale, log = log)
}

suppressWarnings(MASS::fitdistr(x, densfun = dfancy, start = list(shape = 10, scale = 20)))
```


The **stats4**  package provides MLE using S4 classes.
```{r mle}
nll <- function(shape, scale) -sum(dfancy(x, shape, scale, TRUE))
suppressWarnings(fit <- stats4::mle(nll, start = list(shape = 10, scale = 10)))
stats4::summary(fit)
```

## Computer Arithmetics

@burns2012r summarizes many traps beginners may fall into. The first one is the floating number trap. Are you surprised by what you see in the following?
```{r trap}
## floating point traps
.1 == .3 / 3

seq(0, 1, by = .1) == 0.3
seq(0, 1, by = .1)

unique(c(.3, .4 - .1, .5 - .2, .6 - .3, .7 - .4))
c(.3, .4 - .1, .5 - .2, .6 - .3, .7 - .4)

options(digits = 20)
.1
.3 / 3
```

How does a computer represent a number? Computers use switches, which has only two states, on and off, to represent numbers. That is, it uses the binary number system. With a finite (though large) number switches on a computer, not all real numbers are representable. Binary numbers are made of bits, each of which represent a switch. A byte is 8 bits, which represents one character.

### Integers
An integer on many computers has length 4-byte or 32-bit.
It is represented by the coefficients $x_i$ in
\[
\sum_{i=1}^{32} x_i 2^{i-1} - 2^{31},
\]
where $x_i$ is either 0 or 1.

```{r int}
imax <- .Machine$integer.max
imax
imax + 1L
log(imax, base = 2)

u <- 0L
b <- 1L
for (i in 1L:31L) {
  u <- u + b
  b <- b * 2L
}
u

1000L * 1000L * 1000L
1000L * 1000L * 1000L * 1000L
1000L * 1000L * 1000L * 1000
```

### Floating Point
A floating point number is represented by 
\[
(-1)^{x_0}\left(1 + \sum_{i=1}^t x_i 2^{-i}\right) 2^{k}
\]
for $x_i \in \{0, 1\}$,
where $k$ is an integer called the exponent,
$x_0$ is the sign bit.
The fractional part $\sum_{i=1}^t x_i 2^{-i}$ is the significand. By convention (for unique representation), the exponent is chosen so that the fist digit of the significand is 1, unless it would result in the exponent being out of range.


A standard double precision representation uses 8 bytes or 64 bits: a sign bit, an 11 bit exponent, and $t = 52$ bits for the significand. With 11 bits there are $2^{11} = 2048$ possible values for the exponent, which are usually shifted so that the allowed values range from $-1022$ to 1023, again with one special sign value.


The IEEE standard for floating point arithmetic calls for basic arithmetic operations to be performed to higher precision, and then rounded to the nearest representable floating point number. 


Arithmetic underflow occurs where the result of a calculation is a smaller number than what the computer can actually represent. Arithmetic overflow occurs when a calculation produces a result that is greater in magnitude than what the computer can represent. Here is a demonstration [@gray2001advanced].
```{r float}
options(digits = 20)
.Machine$double.eps
.Machine$double.base ^ .Machine$double.ulp.digits
1 + .Machine$double.eps == 1
1 + .Machine$double.base ^ (.Machine$double.ulp.digits - 1L) == 1

## largest possible significand
u <- 0
for (i in 1L:53) u <- u + 2^(-i)
u

## and the largest possible exponent--note that calculating 2^1024
## directly overflows
u * 2 * 2 ^ 1023

## next largest floating point number overflows
(u * 2 * 2 ^ 1023) * (1 + 1/2^52)

## smallest possible mantissa and smallest possible exponent
1 / 2^52 * 2^(-1022)
1 / 2^52 * 2^(-1022) / 2
```

### Error Analysis

More generally, a floating point number with base $\beta$ can be represented as
\[
(-1)^{x_0}\left(\sum_{i=1}^t x_i \beta^{-i}\right) \beta^{k},
\]
where $x_0 \in \{0, 1\}$, $x_i \in \{0, 1, \ldots, \beta - 1\}$, $i = 1, \ldots, t$, and $E_{\min} \le k \le E_{\max}$ for some $E_{\min}$ and $E_{\max}$. For a binary system $\beta = 2$ while for the decimal system $\beta = 10$. To avoid multiple representations, a normalized representation is defined by forcing the first significant digit ($x_1$) to be non-zero. The numbers $\{x: |x| < \beta^{E_{\min} - 1}\}$ (called subnormals) can not be normalized and, are sometimes excluded from the system. If an operation results in a subnormal number underflow is said to have occurred. Similarly if an operation results in a number with the exponent $k > E_{\max}$, overflow is said to have occurred.


Let $f(x)$ be the floating point representation of $x$.
The relative error of $f(x)$ is $\delta_x = [f(x) - x] / x$. The basic formula for error analysis is
\[
f(x \circ y) = (1+ \epsilon_m) (x \circ y) 
\]
where $\circ$ is an operator (e.g., add, subtract, multiple, divide, etc.), $(x \circ y)$ is the exact result, and $\epsilon_m$ is the smallest positive floating-point number $z$ such that $1 + z \ne 1$. __Add demonstration from Section 1.3 of @gray2001advanced; see also [notes] (https://people.eecs.berkeley.edu/~demmel/cs267/lecture21/lecture21.html). __



(Section 1.4 from Gray)
Consider computing $x + y$, where $x$ and $y$ have the same sign. Let $\delta_x$ and $\delta_y$ be the relative error in the floating point representation of $x$ and $y$, respectively (e.g., $\delta_x = [f(x) - x]/x$, so $f(x) = x(1 + \delta_x)$). What the computer actually calculates is the sum of the floating point representations $f (x) + f (y)$, which may not have an exact floating point representation, in which case it is rounded to the nearest representable number $f(f(x) + f(y))$. Let $\delta_s$ be the relative error of the $f(f(x) + f(y))$. Then,
\begin{align*}
&\,\, |f(f(x)+f(y)) - x+y)| \\ 
&= |f(f(x)+f(y)) - f(x) - f(y) + f(x)-x+f(y)-y| \\
&\le |\delta_s[f(x) + f(y)]| + |\delta_x x| + |\delta_y y|\\
&\le |x+y|(\epsilon_m + \epsilon^2_m) + |x+y| \epsilon_m \\
&\approx 2\epsilon_m |x+y|,
\end{align*}
where the higher order terms in $\epsilon_m$ have been dropped, since they are usually negligible. Thus $2\epsilon_m$ is an (approximate) bound on the relative error in a single addition of two numbers with the same sign.

```{example sum, name = "Computing sample sum"}
(Section 1.5.1 of Srivistava (2009))
Let $x_1, \ldots, x_n$ be the observations from a random of size $n$. The simplest algorithm to compute the sample sum is to start from $s_0 = 0$ and add one term each time $s_j = s_{j-1} + x_j$. Each addition may result in a number that is not a floating number, with an error denoted by multiple $(1 + \epsilon)$ to that number. 

For example, when $n = 3$, suppose that $(x_1, x_2, x_3)$ are floating numbers. We have:
\begin{align*}  
f(s_1) &= (0 + x_1)(1 + \epsilon_1)\\
f(s_2) &= (f(s_1) + x_2)(1 + \epsilon_2)\\
f(s_3) &= (f(s_2) + x_3)(1 + \epsilon_3)
\end{align*}
It can be seen that
\begin{align*}
f(s_3) = (x_1 + x_2 + x_3) + x_1 (\epsilon_1 + \epsilon_2 + \epsilon_3) + x_2 (\epsilon_2 + \epsilon_3) + x_3(\epsilon_3) + o(\epsilon_m).
\end{align*}
The cumulative error is approximately
\begin{align*}
&\phantom{=} \| f(s_3) - (x_1 + x_2 + x_3)\| \\
&\sim \| x_1 (\epsilon_1 + \epsilon_2 + \epsilon_3) + x_2 (\epsilon_2 + \epsilon_3) + x_3(\epsilon_3) \|\\
&\le \|x_1\| (\|\epsilon_1\| + \|\epsilon_2\| + \|\epsilon_3\|) + \|x_2\| (\|\epsilon_2\| + \|\epsilon_3\|) + \|x_3\| \|\epsilon_3\| \\
&\le (3 \|x_1\| + 2\|x_2\| + \|x_3\|) \epsilon_m.
\end{align*}
This suggests that better accuracy can be achieved at the expense of sorting in increasing order before computing the sum.
```

__Todo: design a decimal system with limited precision to demonstrate the error analysis __

Subtracting one number from another number of similar magnitude can lead to
large relative error. This is more easily illustrated using a decimal system
($\beta = 10$); see Section~1.4 of @srivastava2009computational. For simplicity, consider the case with $t = 4$. Let $x = 122.9572$ and $y = 123.1498$. Their floating point representations are `(1230, 3)` and `(1231, 3)`. The resulting difference is $-0.1$ while the actual answer is $-0.1926$. The idea is the same on a binary system. 

Consider an extreme case:
\begin{align*}
f(x) &= 1 / 2 + \sum_{i=2}^{t-1} x_i / 2^i + 1 / 2^t,\\
f(y) &= 1 / 2 + \sum_{i=2}^{t-1} x_i / 2^i + 0 / 2^t.
\end{align*}
The absolute errors of $f(x)$ and $f(y)$ are in $(0, 1/2^{t+1})$.
The true difference $x - y$ could be anywhere in $(0, 1 / 2^{t-1})$
The computed subtraction is $f(x) - f(y) = 1 / 2^t$.
So the relative error can be arbitrarily large.


The error from subtraction operation also explains why approximating $e^{x}$ by partial sums does not work well for negative $x$ with a large magnitude. Consider an implementation of the exponential function with the Taylor series
$\exp(x) = \sum_{i=0}^{\infty} x^i / i!$.
```{r fexp}
fexp <- function(x) {
  i <- 0
  expx <- 1
  u <- 1
  while (abs(u) > 1.e-8 * abs(expx)) {
    i <- i + 1
    u <- u * x / i
    expx <- expx + u
  }
  expx
}


options(digits = 10)
x <- c(10, 20, 30)
cbind(exp( x), sapply( x, fexp))
cbind(exp(-x), sapply(-x, fexp))
```

The accuracy is poor for $x < 0$. The problem in accuracy occurs because the terms alternate in sign, and some of the terms are much greater than the final answer. It can be fixed by noting that $\exp(-x) = 1 / \exp(x)$.


```{r fexp2}
fexp2 <- function(x) {
  if (x >= 0) fexp(x) else 1 / fexp(-x)
}

cbind(exp(-x), sapply(-x, fexp2))
```

### Condition Number
(Wiki)
In numerical analysis, the condition number of a function with respect to an argument measures how much the output value of the function can change for a small change in the input argument. This is used to measure how sensitive a function is to changes or errors in the input, and how much error in the output results from an error in the input. 


The condition number is an application of the derivative, and is formally defined as the value of the asymptotic worst-case relative change in output for a relative change in input. The ``function'' is the solution of a problem and the ``arguments'' are the data in the problem. The condition number is frequently applied to questions in linear algebra, in which case the derivative is straightforward but the error could be in many different directions, and is thus computed from the geometry of the matrix. More generally, condition numbers can be defined for non-linear functions in several variables.


A problem with a low condition number is said to be well-conditioned, while a problem with a high condition number is said to be ill-conditioned. The condition number is a property of the problem. Paired with the problem are any number of algorithms that can be used to solve the problem, that is, to calculate the solution. Some algorithms have a property called backward stability. In general, a backward stable algorithm can be expected to accurately solve well-conditioned problems. Numerical analysis textbooks give formulas for the condition numbers of problems and identify the backward stable algorithms. In the exponential function example, the first implementation is unstable for large negative $x$ while the second is stable.


It is important in likelihood maximization to work on the log scale. Use argument `log` in `dpq` functions instead of taking logs. This can be important in probalistic networks and MC(MC) where $P = P_1 · P_2 \cdots  · P_n$ quickly underflows to zero. 


It is also important to use `log1p` and `expm1` alike when caculating $\log(1 + x)$ and $\exp(x)- 1$  when $|x| \ll 1$. One example is the evaluation of distribution and density function of the generalized extreme value (GEV) distribution. The GEV distribution has distribution function
\begin{equation*}
F(x; \mu, \sigma, \xi) = 
\begin{cases}
\exp[- \left[ 1 + \left(\frac{x - \mu}{\sigma}\right) \xi\right]^{-1 / \xi} & \xi \ne 0, \quad (1 + \xi (x - \mu) / \sigma) > 0,\\
\exp[e^{ -(x - \mu) / \sigma}] & \xi = 0,
\end{cases}
\end{equation*}
for $\mu \in R$, $\sigma > 0$, $\xi \in R$.
```{r gev, fig.asp = 1, out.width='50%', fig.show='hold', fig.cap='Two implementations of GEV distribution function when $\\xi$ is close to zero.'}
curve(evir::pgev(1, xi = x),    1e-20, .01, log="x", n = 1025)
curve(texmex::pgev(1, 0, 1, x), 1e-20, .01, log="x", n = 1025)
```
Figure \@ref(fig:gev) shows a comparison of evaluating the GEV distribution function at $x = 1$ for $\mu = 0$, $\sigma = 1$, and a sequence of $\xi$ values near zero using two implementations.


## Exercises
1. Use git to clone the source of Hadley Wickham's **Advanced R Programming** from his GitHub repository to a local space on your own computer. Build the book using RStudio. During the building process, you may see error messages due to missing tools on your computer. Read the error messages carefully and fix them, until you get the book built. You may need to install some R packages, some fonts, some latex packages, and some building tools for R packages. On Windows, some codes for parallel computing may not work and need to be swapped out. Document the problems you encountered and how you solved them in an R Markdown file named `README.Rmd`. Push it to your homework GitHub repository so that it can help other students to build the book. 


2. Use **bookdown** or **rmarkdown** to produce a report for the following task. Consider approximation of the distribution function of $N(0, 1)$,
\begin{equation}
\Phi(t) = \int_{-\infty}^t \frac{1}{\sqrt{2\pi}} e^{-y^2 / 2} dy,
(\#eq:cdf)
\end{equation}
by the Monte Carlo methods:
\begin{equation}
\hat\Phi(t) = \frac{1}{n} \sum_{i=1}^n I(X_i \le t),
\end{equation}
where $X_i$'s are iid $N(0, 1)$ variables.
Experiment with the approximation at $n \in \{10^2, 10^3, 10^4\}$ at $t \in \{0.0, 0.67, 0.84, 1.28, 1.65, 2.32, 2.58, 3.09, 3.72\}$ to form a table.


The table should include the true value for comparison.
Further, repeat the experiment 100 times. Draw box plots of the bias at all $t$.
The report should look like a manuscript, with a title, an abstract, and multiple sections. It should contain at least one math equation, one table, one figure, and one chunk of R code. The template of our Data Science Lab can be helpful: <https://statds.org/template/>, the source of which is at <https://github.com/statds/dslab-templates>.


3. Explain how `.Machine$double.xmax`, `.Machine$double.xmin`, `.Machine$double.eps`, and `.Machine@double.neg.eps` are defined using the 64-bit double precision floating point arithmetic.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown',
  "MASS"
), 'packages.bib')
```
